---
title: "final-code"
format: 
  html: 
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
editor: visual
execute: 
  message: false
  warning: false
---

## Final

# Problem 1: Kangaroo rat seed banks

# Introduction

Kangaroo rats are small, mostly nocturnal rodents that are native to western North America in arid areas. The mounds they create can result in the formation of microhabitats that support unique plant communities and local species diversity (Guo 1). This in part is due to the fact that these mounds possess more organic matter to promote plant growth (Davidson and Lightfoot 1). Although there is evidence of more plant growth on these mounds, there is not much data on the composition of the seed bank on kangaroo rat mounds (Moorhead et.al). Therefore this analysis serves to understand how total seed number differs between kangaroo rat mound locations in a Chihuahuan Desert Grassland. We are testing the null hypothesis that seed numbers do not different between kangaroo rat mound locations against the alternative hypothesis that seed numbers do differ between kangaroo rat mound locations. 

# Methods

The data from the dataset was compiled by conducting an experiment in which 10 kangaroo rat mounds were randomly selected to collect soil samples from. In addition, subsamples were taken from each mound and grouped into 4 main categories: base of the mound (base), one meter from base samples (surrounding), six meters from base samples at the edge of black grama glass clumps (edge), and six meters from base samples in the interspace between black grama grass clumps (interspace). Soil samples were then dried in an oven for 48 hours at 50 C before being sifted using a sieve to capture small seeds. Eight taxa of seeds were targeted: Cryptantha crassisepala, Descurainia pinnata, Phacelia integrifolia, and Plantago patagonica (spring annuals), Astragalus missouriensis, Lesquerella fendleri, and Oenothera species (perrenial forbs that flower in the spring), and Sphaeralacea species (perennail forb that flowers in the fall). To organize these data, the column names were renamed to be more informative using the rename() function. 

```{r set-up}
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(performance)
library(naniar)
library(flextable)
library(car)
library(broom)
library(ggplot2)
library(corrplot)
library(AICcmodavg)
library(GGally)
library(MuMIn)
```

Read in the data:

```{r}
kangaroo_data <- read_csv(here("data", "kangaroo_mound_data.csv")) %>% 
  # rename columns to be more informative
  rename(c("mound" = "mnd", "direction" = "dir", "microhabitat" = "loc", "seed_count" = "seeds"))
```

Next, a visualization of the missing data were created which showed there were no missing data.

```{r}
gg_miss_var(kangaroo_data)
```
**Figure 1: Missing Data.** Number of missing data entires on the x-axis and variables on the y-axis. Plots correspond to number of missing data entries for each variable. 

To determine whether seed numbers differ between kangaroo rat mound locations an ANOVA test was conducted. First the assumptions for this test were checked including normality and variance. 

```{r}
ggplot(data = kangaroo_data) +
  stat_qq(aes(sample = seed_count)) +
  stat_qq_line(aes(sample = seed_count), color = "blue") +
  facet_wrap(~ mound)
```


```{r}
count_var <- leveneTest(seed_count ~ mound, data = kangaroo_data)

count_var
```













# Works Cited 

https://link.springer.com/article/10.1007/bf00328605?utm_source=getftr&utm_medium=getftr&utm_campaign=getftr_pilot

https://www.jstor.org/stable/2426018?origin=crossref

https://pdf.sciencedirectassets.com/272559/1-s2.0-S0140196308X00059/1-s2.0-S014019630700345X/main.pdf?X-Amz-Security-Token=IQoJb3JpZ2luX2VjEAUaCXVzLWVhc3QtMSJIMEYCIQCJhda6E0FSmJMmYJVWyUomKv%2Bcho1LLFrAVIRRcJNG1AIhAMut05%2BlEIWjO3DWs1pCXxSVgMDKJzKyMvNRD66%2BljYjKrIFCE0QBRoMMDU5MDAzNTQ2ODY1IgylYW%2FwtWIv5JWOzncqjwW%2F7gFInPENcKagkFSJi9UimxEwuy0RKODPF0bZN%2BzvyWQ4T3md4TbBd2yL4YSU%2FKZQIj4UscIcSVtEX51dAXsymSKHgKd5Vf95eEXunqg84asKx3kSAUt0z%2BLuaasbaxdaMU7Ym0gIrjf%2Fo9rhVJ385PEv3V1PR4dfhzfnN%2BxnjHtouvRUSplIawkYcmzgCuodo%2B%2B27EqfKEFf2yw8D9NxT7zzbK7puAojSzRBooH7d28JcvZfkyvqFukUcoaxq9h87qAkjvL5znzpfRBFDRQbcON2ALvaaJAzquLBatvAPR%2FYtTC1AzhCUg2e1z3DB9cGI3u%2BjFVI2ueEs0n%2B5G3aYFs62TOxA1hRUIVXYY8jjEWpsmb8Qll5BieaDXB%2BnFvG%2FQ%2F2ko6aZtX8SdBLXa8HBbCDKOXyBAGg8Q0%2BAhEWcRIjp%2BVx%2FxzbLvHqA8%2ByopDQpEfYWGCwr12uZqoLVKHq2eM0mS%2BapFs7LWkCh8LA1v7NLn1bUOiR3vtAFv4r5ROynyP28PW0BFvTBj5jNds4tvPx0gSByoHBhGBv7U4ldgtlQCdbhaB4NnkCM1LDZ9IeCw%2BMlVwF3sm6Jt%2B0BOnLPcNhR99aNbO2ilSA2tQd1gQwbFCoCCkWkBVV0Uf%2Ff7Puk2Af5IcwH24JfUd%2FCz8aNa%2FyTMeqBGCeNhU4oLrY12QQN5vbuBMfwO1njn1UBUsFnr%2Fm%2BaQK35TsAw7jKAygULP9NOEpXo%2BoVGBKHh8PEv1gnp26txpnoczA%2BbiwtPMNps7dAgOw77gKCp9pjKA4OiUuYS6iZHoW1cucW%2FNLHM7q6gS0nCH8vqJh5XAp%2FuWIUOqL%2FcZk11X0HllrdBxxPpnp8NxLUrJqI9RZ5GoZMP%2Bk%2FqMGOrABRVsprCDl5PXeIPp3xCZsNJnXNdN2iI5%2Fj%2BoqzPOaUAXL37nRBpR9BMKWp7hg6rYGr4joWSNBi2bU0cqW%2BffYhrmbBRS1d%2BC5xZ6bRv%2F8J6z%2FfLL0yu169kyHmzTh0I9hzxCNTlv6TGDuAo%2BpSl4ljLwneHEsC8D0QfQ8Qmgtn31pRhP7eyWq6%2FxMKP4uhUIQpiW%2BidgSBUuTAmcYYqjTIRYOhLSHkQSPxno%2B1zXqb8g%3D&X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Date=20230606T213756Z&X-Amz-SignedHeaders=host&X-Amz-Expires=300&X-Amz-Credential=ASIAQ3PHCVTY776KDSP3%2F20230606%2Fus-east-1%2Fs3%2Faws4_request&X-Amz-Signature=18f7de3fcb18a1265bed0c8d9a10709f067efc7f345b11c87e92c563eec6f561&hash=74952f99329839540e93b032b1f0f2ba2c244a69e00c7bc5ea35d035a4c104df&host=68042c943591013ac2b2430a89b270f6af2c76d8dfd086a07176afe7c76c2c61&pii=S014019630700345X&tid=spdf-9d03e12e-c97a-4c53-b743-4f0673e3e4c7&sid=9308c8849183a247132a37f-d41f656897c3gxrqa&type=client&tsoh=d3d3LnNjaWVuY2VkaXJlY3QuY29t&ua=0f1c560b5c5d05050052&rr=7d33cca4ab567d5c&cc=us